---
title: "run_Horvarth_model"
author: "Rogério Ribeiro"
date: "`r Sys.Date()`"
output: html_document
---

# Predict age using the Horvath model


## Load libraries
```{r}
suppressPackageStartupMessages(library(tidyverse))
library(readr)
library(lmtest)
library(data.table)
library(impute)
library(Metrics)
library(MLmetrics)
```

```{r}
source("normalization.R")
```


```{r}
safe_colorblind_palette <- c("#88CCEE", "#CC6677", "#DDCC77", "#117733", "#332288", "#AA4499", 
                             "#44AA99", "#999933", "#882255", "#661100", "#6699CC", "#888888")

```

```{r}
# Replace this folder
image_folder <- "../../aging_notes/figures/2.epigenetic_clocks/"
```



## Function and data
```{r}
trafo= function(x,adult.age=20) { x=(x+1)/(1+adult.age); y=ifelse(x<=1, log( x),x-1);y }
anti.trafo= function(x,adult.age=20) { ifelse(x<0, (1+adult.age)*exp(x)-1, (1+adult.age)*x+adult.age) }
probeAnnotation21kdatMethUsed=read.csv("data/probeAnnotation21kdatMethUsed.csv")
probeAnnotation27k=read.csv("data/datMiniAnnotation27k.csv")
datClock=read.csv("data/AdditionalFile3.csv")
```

## Lung

### Load my data and data preparation
```{r}
# Load methylation data
lung <- fread(file = "../../data/methylation_lung.csv")

#Load metadata 
metadata <- read.csv(file = "../../metadata/eGTExDNA_Pierce_Jan18.09-11-2021.tsv", sep = "\t")
metadata <- metadata %>% filter(`Tissue.Site.Detail` == "Lung")
age_data <- read.csv(file = "../../metadata/lung_annotation_meth.csv")


lung_test_samples <- read.csv(file = "../../metadata/lung_test_metadata.csv")


#Filter for samples with age data
samples_with_age <- age_data %>% pull(Sample_ID)
lung <- lung %>% 
  select(probe, all_of(samples_with_age))

```

Keep only probes in the 21K (we have some missing values, which we will set to NA)

```{r}
lung <- lung %>% 
  filter(probe %in% probeAnnotation21kdatMethUsed$Name)


match1 <- match(probeAnnotation21kdatMethUsed$Name , lung$probe)
sum(is.na(match1))
```

Due to missing genes from the 21K probe array, we need to input them with NA values

```{r}
missing_probes <- probeAnnotation21kdatMethUsed$Name[is.na(match1)]
missing_probes_data <- matrix(data = NA, nrow = length(missing_probes), ncol = ncol(lung))
missing_probes_data[,1] <- missing_probes
colnames(missing_probes_data) <-  colnames(lung)

missing_probes_data <- as.data.frame(missing_probes_data)

lung <- rbind(lung, missing_probes_data)
lung <- lung %>% 
  as_tibble() %>%
  mutate_at(2:ncol(lung), as.numeric)
  
```

```{r}
print(dim(lung))
```

Sanity check for missing data

```{r}

match2 <- match(probeAnnotation21kdatMethUsed$Name , lung$probe)
sum(is.na(match2))
```

### Script based on Horvath "step-wise analysis"

#### STEP 1: DEFINE QUALITY METRICS


```{r}

meanMethBySample = as.numeric(apply(as.matrix(lung[,-1]),2,mean,na.rm=TRUE))
minMethBySample = as.numeric(apply(as.matrix(lung[,-1]),2,min,na.rm=TRUE))
maxMethBySample = as.numeric(apply(as.matrix(lung[,-1]),2,max,na.rm=TRUE))

datMethUsed= t(lung[,-1])
colnames(datMethUsed)=as.character(lung$probe)


noMissingPerSample=apply(as.matrix(is.na(datMethUsed)),1,sum)
table(noMissingPerSample)

```

#### STEP 2: Imputing and normalization

Run the following code if there is at least one missing

```{r}
if (max(noMissingPerSample,na.rm=TRUE)<3000 ){
  if ( max(noMissingPerSample,na.rm=TRUE)>0 ){
    dimnames1=dimnames(datMethUsed)
    datMethUsed= data.frame(t(impute.knn(t(datMethUsed))$data))
    dimnames(datMethUsed)=dimnames1
  } # end of if
} # end of if (! fastImputation )

#Note There another faster imputation method in the script


```

#### Step 2.5 Normalize the data

```{r include=FALSE}
source("normalization.R")

datMethUsedNormalized = BMIQcalibration(
  datM = datMethUsed,
  goldstandard.beta = probeAnnotation21kdatMethUsed$goldstandard2,
  plots = FALSE
)
```


#### Step 3: Split the data into train and test

```{r}
split_in_train_test <- function(meth, age_data, test_set) {
  # Metadata
  metadata_test <- age_data[age_data$tissue_sample_id %in% test_set$sample_id, ]
  metadata_train <- age_data[!age_data$tissue_sample_id %in% test_set$sample_id, ]

  # Age data
  y_test <- metadata_test$AGE
  y_train <- metadata_train$AGE

  # Methylation data
  X_train <- meth[metadata_train$Sample_ID, ]
  X_test <- meth[metadata_test$Sample_ID, ]

  return(list(X_train, X_test, y_train, y_test))
}


train_test <- split_in_train_test(datMethUsedNormalized, age_data, lung_test_samples)

X_train <- train_test[[1]]
X_test <- train_test[[2]]

y_train <- train_test[[3]]
y_test <- train_test[[4]]

```

#### STEP 4: Predict age and create a data frame for the output 

```{r}

selectCpGsClock <- is.element(dimnames(datMethUsed)[[2]], as.character(datClock$CpGmarker[-1]))

## Train
datMethClock0 <- data.frame(X_train[,selectCpGsClock])
datMethClock <-  data.frame(datMethClock0[ as.character(datClock$CpGmarker[-1])])
predictedAge_lung_train <- as.numeric(anti.trafo(datClock$CoefficientTraining[1]+as.matrix(datMethClock)%*% as.numeric(datClock$CoefficientTraining[-1])))

## Test
datMethClock0 <- data.frame(X_test[,selectCpGsClock])
datMethClock <-  data.frame(datMethClock0[ as.character(datClock$CpGmarker[-1])])
predictedAge_lung_test <- as.numeric(anti.trafo(datClock$CoefficientTraining[1]+as.matrix(datMethClock)%*% as.numeric(datClock$CoefficientTraining[-1])))

```

#### Step 5: Compute accuracy metrics

In train data

```{r}

# Calculate Mean Absolute Error (MAE)
mae_value <- mae(predictedAge_lung_train, y_train)

# Calculate Mean Squared Error (MSE)
mse_value <- mse(predictedAge_lung_train, y_train)

# Calculate Root Mean Squared Error (RMSE)
rmse_value <- sqrt(mse_value)

# Calculate median absolute error
med <- median(abs(predictedAge_lung_train - y_train))

# Compute correlation
cor <- cor(predictedAge_lung_train, y_train)

# Calculate R-squared (R²)
rsquared_value <- R2_Score(predictedAge_lung_train, y_train)

metrics_horvarth_lung_train <- c("Lung", "train", mae_value, med, rmse_value, rsquared_value, cor)

# Plot results
results_lung_train <- data.frame("predicted" = predictedAge_lung_train, "age" = y_train, "sample" =  row.names(X_train))


a <- ggplot(results_lung_train, aes(x = age, y = predicted)) +
  geom_point() +  # Add scatter points
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +  # Add a 45-degree line
  #annotate(paste("MAE =", round(mae_value, 2), "\n", "RMSE =", round(rmse_value, 2))) + 
  annotate("text", x=6, y=60, label = paste("MAE =", round(mae_value, 2), "\n", "MED =", round(med, 2), "\nRMSE =", round(rmse_value, 2), "\nR2 =", round(rsquared_value, 2), "\ncor =", round(cor,2))) + 
  labs(x = "Real Age", y = "Predicted Age") +  # Label axes
  ggtitle(paste0("Real Age vs. Predicted Age (Train set, n = ", nrow(results_lung_train),")")) +  # Add a title
  theme_classic() + 
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size = 18)) + 
  xlim(0, max(results_lung_train$age)) +
  ylim(0, max(results_lung_train$predicted))


print(a)

# Create a scatter plot of real age vs. predicted age
png(paste0(image_folder, "2.horvarth_lung_train_real_vs_predicted.png"), res = 300, w = 2000, h = 1500)
a
dev.off()

```

In test data

```{r}
# Calculate Mean Absolute Error (MAE)
mae_value <- mae(predictedAge_lung_test, y_test)

# Calculate Mean Squared Error (MSE)
mse_value <- mse(predictedAge_lung_test, y_test)

# Calculate Root Mean Squared Error (RMSE)
rmse_value <- sqrt(mse_value)

# Calculate median absolute error
med <- median(abs(predictedAge_lung_test - y_test))

# Compute correlation
cor <- cor(predictedAge_lung_test, y_test)

# Calculate R-squared (R²)
rsquared_value <- R2_Score(predictedAge_lung_test, y_test)

metrics_horvarth_lung_test <- c("Lung", "test", mae_value, med, rmse_value, rsquared_value, cor)

# Plot results
results_lung_test <- data.frame("predicted" = predictedAge_lung_test, "age" = y_test, "sample" =  row.names(X_test))


a <- ggplot(results_lung_test, aes(x = age, y = predicted)) +
  geom_point() +  # Add scatter points
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +  # Add a 45-degree line
  #annotate(paste("MAE =", round(mae_value, 2), "\n", "RMSE =", round(rmse_value, 2))) + 
  annotate("text", x=6, y=60, label = paste("MAE =", round(mae_value, 2), "\n", "MED =", round(med, 2), "\nRMSE =", round(rmse_value, 2), "\nR2 =", round(rsquared_value, 2), "\ncor =", round(cor,2))) + 
  labs(x = "Real Age", y = "Predicted Age") +  # Label axes
  ggtitle(paste0("Real Age vs. Predicted Age (Test set, n = ", nrow(results_lung_test),")")) +  # Add a title
  theme_classic() + 
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size = 18)) + 
  xlim(0, max(results_lung_test$age)) +
  ylim(0, max(results_lung_test$predicted))


print(a)

# Create a scatter plot of real age vs. predicted age
png(paste0(image_folder, "2.horvarth_lung_test_real_vs_predicted.png"), res = 300, w = 2000, h = 1500)
a
dev.off()

```

```{r}
results_lung_test
```


## Colon - Transverse

### Load my data and data preparation
```{r}
# Load methylation data
colon <- fread(file = "../../data/methylation_colon.csv")

#Load metadata 
metadata <- read.csv(file = "../../metadata/eGTExDNA_Pierce_Jan18.09-11-2021.tsv", sep = "\t")
metadata <- metadata %>% filter(`Tissue.Site.Detail` == "Colon - Transverse")
age_data <- read.csv(file = "../../metadata/colon_annotation_meth.csv")

colon_test_samples <- read.csv(file = "../../metadata/colon_test_metadata.csv")


#Filter for samples with age data
samples_with_age <- age_data %>% pull(Sample_ID)
colon <- colon %>% 
  select(probe, all_of(samples_with_age))

```

Keep only probes in the 21K (we have some missing values, which we will set to NA)

```{r}
colon <- colon %>% 
  filter(probe %in% probeAnnotation21kdatMethUsed$Name)


match1 <- match(probeAnnotation21kdatMethUsed$Name , colon$probe)
sum(is.na(match1))
```

Due to missing genes from the 21K probe array, we need to input them with NA values

```{r}
missing_probes <- probeAnnotation21kdatMethUsed$Name[is.na(match1)]
missing_probes_data <- matrix(data = NA, nrow = length(missing_probes), ncol = ncol(colon))
missing_probes_data[,1] <- missing_probes
colnames(missing_probes_data) <-  colnames(colon)

missing_probes_data <- as.data.frame(missing_probes_data)

colon <- rbind(colon, missing_probes_data)
colon <- colon %>% 
  as_tibble() %>%
  mutate_at(2:ncol(colon), as.numeric)
  
```

```{r}
print(dim(colon))
```

Sanity check for missing data

```{r}

match2 <- match(probeAnnotation21kdatMethUsed$Name , colon$probe)
sum(is.na(match2))
```

### Script based on Horvath "step-wise analysis"

#### STEP 1: DEFINE QUALITY METRICS


```{r}

meanMethBySample = as.numeric(apply(as.matrix(colon[,-1]),2,mean,na.rm=TRUE))
minMethBySample = as.numeric(apply(as.matrix(colon[,-1]),2,min,na.rm=TRUE))
maxMethBySample = as.numeric(apply(as.matrix(colon[,-1]),2,max,na.rm=TRUE))

datMethUsed= t(colon[,-1])
colnames(datMethUsed)=as.character(colon$probe)


noMissingPerSample=apply(as.matrix(is.na(datMethUsed)),1,sum)
table(noMissingPerSample)

```

#### STEP 2: Imputing 

Run the following code if there is at least one missing

```{r include=FALSE}
if (max(noMissingPerSample,na.rm=TRUE)<3000 ){
  if ( max(noMissingPerSample,na.rm=TRUE)>0 ){
    dimnames1=dimnames(datMethUsed)
    datMethUsed= data.frame(t(impute.knn(t(datMethUsed))$data))
    dimnames(datMethUsed)=dimnames1
  } # end of if
} # end of if (! fastImputation )

#Note There another faster imputation method in the script


```


#### Step 2.5 Normalize the data

```{r include=FALSE}
source("normalization.R")

datMethUsedNormalized = BMIQcalibration(
  datM = datMethUsed,
  goldstandard.beta = probeAnnotation21kdatMethUsed$goldstandard2,
  plots = FALSE
)
```

#### Step 3: Split the data into train and test


```{r}
split_in_train_test <- function(meth, age_data, test_set) {
  # Metadata
  metadata_test <- age_data[age_data$tissue_sample_id %in% test_set$sample_id, ]
  metadata_train <- age_data[!age_data$tissue_sample_id %in% test_set$sample_id, ]

  # Age data
  y_test <- metadata_test$AGE
  y_train <- metadata_train$AGE

  # Methylation data
  X_train <- meth[metadata_train$Sample_ID, ]
  X_test <- meth[metadata_test$Sample_ID, ]

  return(list(X_train, X_test, y_train, y_test))
}


train_test <- split_in_train_test(datMethUsedNormalized, age_data, colon_test_samples)

X_train <- train_test[[1]]
X_test <- train_test[[2]]

y_train <- train_test[[3]]
y_test <- train_test[[4]]

```

#### STEP 4: Predict age and create a data frame for the output 

```{r}

selectCpGsClock <- is.element(dimnames(datMethUsed)[[2]], as.character(datClock$CpGmarker[-1]))

## Train
datMethClock0 <- data.frame(X_train[,selectCpGsClock])
datMethClock <-  data.frame(datMethClock0[ as.character(datClock$CpGmarker[-1])])
predictedAge_colon_train <- as.numeric(anti.trafo(datClock$CoefficientTraining[1]+as.matrix(datMethClock)%*% as.numeric(datClock$CoefficientTraining[-1])))

## Test
datMethClock0 <- data.frame(X_test[,selectCpGsClock])
datMethClock <-  data.frame(datMethClock0[ as.character(datClock$CpGmarker[-1])])
predictedAge_colon_test <- as.numeric(anti.trafo(datClock$CoefficientTraining[1]+as.matrix(datMethClock)%*% as.numeric(datClock$CoefficientTraining[-1])))

```

#### Step 5: Compute accuracy metrics

In train data

```{r}

# Calculate Mean Absolute Error (MAE)
mae_value <- mae(predictedAge_colon_train, y_train)

# Calculate Mean Squared Error (MSE)
mse_value <- mse(predictedAge_colon_train, y_train)

# Calculate Root Mean Squared Error (RMSE)
rmse_value <- sqrt(mse_value)

# Calculate median absolute error
med <- median(abs(predictedAge_colon_train - y_train))

# Compute correlation
cor <- cor(predictedAge_colon_train, y_train)

# Calculate R-squared (R²)
rsquared_value <- R2_Score(predictedAge_colon_train, y_train)

metrics_horvarth_colon_train <- c("colon", "train", mae_value, med, rmse_value, rsquared_value, cor)

# Plot results
results_colon_train <- data.frame("predicted" = predictedAge_colon_train, "age" = y_train, "sample" =  row.names(X_train))


a <- ggplot(results_colon_train, aes(x = age, y = predicted)) +
  geom_point() +  # Add scatter points
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +  # Add a 45-degree line
  #annotate(paste("MAE =", round(mae_value, 2), "\n", "RMSE =", round(rmse_value, 2))) + 
  annotate("text", x=6, y=60, label = paste("MAE =", round(mae_value, 2), "\n", "MED =", round(med, 2), "\nRMSE =", round(rmse_value, 2), "\nR2 =", round(rsquared_value, 2), "\ncor =", round(cor,2))) + 
  labs(x = "Real Age", y = "Predicted Age") +  # Label axes
  ggtitle(paste0("Real Age vs. Predicted Age (Train set, n = ", nrow(results_colon_train),")")) +  # Add a title
  theme_classic() + 
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size = 18)) + 
  xlim(0, max(results_colon_train$age)) +
  ylim(0, max(results_colon_train$predicted))


print(a)

# Create a scatter plot of real age vs. predicted age
png(paste0(image_folder, "2.horvarth_colon_train_real_vs_predicted.png"), res = 300, w = 2000, h = 1500)
a
dev.off()

```

In test data

```{r}
# Calculate Mean Absolute Error (MAE)
mae_value <- mae(predictedAge_colon_test, y_test)

# Calculate Mean Squared Error (MSE)
mse_value <- mse(predictedAge_colon_test, y_test)

# Calculate Root Mean Squared Error (RMSE)
rmse_value <- sqrt(mse_value)

# Calculate median absolute error
med <- median(abs(predictedAge_colon_test - y_test))

# Compute correlation
cor <- cor(predictedAge_colon_test, y_test)

# Calculate R-squared (R²)
rsquared_value <- R2_Score(predictedAge_colon_test, y_test)

metrics_horvarth_colon_test <- c("colon", "test", mae_value, med, rmse_value, rsquared_value, cor)

# Plot results
results_colon_test <- data.frame("predicted" = predictedAge_colon_test, "age" = y_test, "sample" =  row.names(X_test))


a <- ggplot(results_colon_test, aes(x = age, y = predicted)) +
  geom_point() +  # Add scatter points
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +  # Add a 45-degree line
  #annotate(paste("MAE =", round(mae_value, 2), "\n", "RMSE =", round(rmse_value, 2))) + 
  annotate("text", x=6, y=60, label = paste("MAE =", round(mae_value, 2), "\n", "MED =", round(med, 2), "\nRMSE =", round(rmse_value, 2), "\nR2 =", round(rsquared_value, 2), "\ncor =", round(cor,2))) + 
  labs(x = "Real Age", y = "Predicted Age") +  # Label axes
 ggtitle(paste0("Real Age vs. Predicted Age (Test set, n = ", nrow(results_colon_test),")")) +  # Add a title
  theme_classic() + 
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size = 18)) + 
  xlim(0, max(results_colon_test$age)) +
  ylim(0, max(results_colon_test$predicted))


print(a)

# Create a scatter plot of real age vs. predicted age
png(paste0(image_folder, "2.horvarth_colon_test_real_vs_predicted.png"), res = 300, w = 2000, h = 1500)
a
dev.off()

```


## Ovary

### Load my data and data preparation
```{r}
# Load methylation data
ovary <- fread(file = "../../data/methylation_ovary.csv")

#Load metadata 
metadata <- read.csv(file = "../../metadata/eGTExDNA_Pierce_Jan18.09-11-2021.tsv", sep = "\t")
metadata <- metadata %>% filter(`Tissue.Site.Detail` == "Ovary")
age_data <- read.csv(file = "../../metadata/ovary_annotation_meth.csv")


ovary_test_samples <- read.csv(file = "../../metadata/ovary_test_metadata.csv")


#Filter for samples with age data
samples_with_age <- age_data %>% pull(Sample_ID)
ovary <- ovary %>% 
  select(probe, all_of(samples_with_age))

```

Keep only probes in the 21K (we have some missing values, which we will set to NA)

```{r}
ovary <- ovary %>% 
  filter(probe %in% probeAnnotation21kdatMethUsed$Name)


match1 <- match(probeAnnotation21kdatMethUsed$Name , ovary$probe)
sum(is.na(match1))
```

Due to missing genes from the 21K probe array, we need to input them with NA values

```{r}
missing_probes <- probeAnnotation21kdatMethUsed$Name[is.na(match1)]
missing_probes_data <- matrix(data = NA, nrow = length(missing_probes), ncol = ncol(ovary))
missing_probes_data[,1] <- missing_probes
colnames(missing_probes_data) <-  colnames(ovary)

missing_probes_data <- as.data.frame(missing_probes_data)

ovary <- rbind(ovary, missing_probes_data)
ovary <- ovary %>% 
  as_tibble() %>%
  mutate_at(2:ncol(ovary), as.numeric)
  
```

```{r}
print(dim(ovary))
```

Sanity check for missing data

```{r}

match2 <- match(probeAnnotation21kdatMethUsed$Name , ovary$probe)
sum(is.na(match2))
```

### Script based on Horvath "step-wise analysis"

#### STEP 1: DEFINE QUALITY METRICS


```{r}
meanMethBySample = as.numeric(apply(as.matrix(ovary[,-1]),2,mean,na.rm=TRUE))
minMethBySample = as.numeric(apply(as.matrix(ovary[,-1]),2,min,na.rm=TRUE))
maxMethBySample = as.numeric(apply(as.matrix(ovary[,-1]),2,max,na.rm=TRUE))

datMethUsed= t(ovary[,-1])
colnames(datMethUsed)=as.character(ovary$probe)


noMissingPerSample=apply(as.matrix(is.na(datMethUsed)),1,sum)
table(noMissingPerSample)

```

#### STEP 2: Imputing 

Run the following code if there is at least one missing

```{r}

if ( max(noMissingPerSample,na.rm=TRUE)>0 ){
  dimnames1=dimnames(datMethUsed)
  datMethUsed= data.frame(t(impute.knn(t(datMethUsed))$data))
  dimnames(datMethUsed)=dimnames1
} # end of if

#Note There another faster imputation method in the script


```


#### Step 2.5 Normalize the data

```{r include=FALSE}
source("normalization.R")

datMethUsedNormalized = BMIQcalibration(
  datM = datMethUsed,
  goldstandard.beta = probeAnnotation21kdatMethUsed$goldstandard2,
  plots = FALSE
)
```


#### Step 3: Split the data into train and test


```{r}
split_in_train_test <- function(meth, age_data, test_set) {
  # Metadata
  metadata_test <- age_data[age_data$tissue_sample_id %in% test_set$sample_id, ]
  metadata_train <- age_data[!age_data$tissue_sample_id %in% test_set$sample_id, ]

  # Age data
  y_test <- metadata_test$AGE
  y_train <- metadata_train$AGE

  # Methylation data
  X_train <- meth[metadata_train$Sample_ID, ]
  X_test <- meth[metadata_test$Sample_ID, ]

  return(list(X_train, X_test, y_train, y_test))
}


train_test <- split_in_train_test(datMethUsedNormalized, age_data, ovary_test_samples)

X_train <- train_test[[1]]
X_test <- train_test[[2]]

y_train <- train_test[[3]]
y_test <- train_test[[4]]

```

#### STEP 4: Predict age and create a data frame for the output 

```{r}

selectCpGsClock <- is.element(dimnames(datMethUsed)[[2]], as.character(datClock$CpGmarker[-1]))

## Train
datMethClock0 <- data.frame(X_train[,selectCpGsClock])
datMethClock <-  data.frame(datMethClock0[ as.character(datClock$CpGmarker[-1])])
predictedAge_ovary_train <- as.numeric(anti.trafo(datClock$CoefficientTraining[1]+as.matrix(datMethClock)%*% as.numeric(datClock$CoefficientTraining[-1])))

## Test
datMethClock0 <- data.frame(X_test[,selectCpGsClock])
datMethClock <-  data.frame(datMethClock0[ as.character(datClock$CpGmarker[-1])])
predictedAge_ovary_test <- as.numeric(anti.trafo(datClock$CoefficientTraining[1]+as.matrix(datMethClock)%*% as.numeric(datClock$CoefficientTraining[-1])))

```

#### Step 5: Compute accuracy metrics

In train data

```{r}

# Calculate Mean Absolute Error (MAE)
mae_value <- mae(predictedAge_ovary_train, y_train)

# Calculate Mean Squared Error (MSE)
mse_value <- mse(predictedAge_ovary_train, y_train)

# Calculate Root Mean Squared Error (RMSE)
rmse_value <- sqrt(mse_value)

# Calculate median absolute error
med <- median(abs(predictedAge_ovary_train - y_train))

# Compute correlation
cor <- cor(predictedAge_ovary_train, y_train)

# Calculate R-squared (R²)
rsquared_value <- R2_Score(predictedAge_ovary_train, y_train)

metrics_horvarth_ovary_train <- c("ovary", "train", mae_value, med, rmse_value, rsquared_value, cor)

# Plot results
results_ovary_train <- data.frame("predicted" = predictedAge_ovary_train, "age" = y_train, "sample" =  row.names(X_train))


a <- ggplot(results_ovary_train, aes(x = age, y = predicted)) +
  geom_point() +  # Add scatter points
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +  # Add a 45-degree line
  #annotate(paste("MAE =", round(mae_value, 2), "\n", "RMSE =", round(rmse_value, 2))) + 
  annotate("text", x=6, y=60, label = paste("MAE =", round(mae_value, 2), "\n", "MED =", round(med, 2), "\nRMSE =", round(rmse_value, 2), "\nR2 =", round(rsquared_value, 2), "\ncor =", round(cor,2))) + 
  labs(x = "Real Age", y = "Predicted Age") +  # Label axes
  ggtitle(paste0("Real Age vs. Predicted Age (Train set, n = ", nrow(results_ovary_train),")")) +  # Add a title
  theme_classic() + 
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size = 18)) + 
  xlim(0, max(results_ovary_train$age)) +
  ylim(0, max(results_ovary_train$predicted))


print(a)

# Create a scatter plot of real age vs. predicted age
png(paste0(image_folder, "2.horvarth_ovary_train_real_vs_predicted.png"), res = 300, w = 2000, h = 1500)
a
dev.off()

```

In test data

```{r}
# Calculate Mean Absolute Error (MAE)
mae_value <- mae(predictedAge_ovary_test, y_test)

# Calculate Mean Squared Error (MSE)
mse_value <- mse(predictedAge_ovary_test, y_test)

# Calculate Root Mean Squared Error (RMSE)
rmse_value <- sqrt(mse_value)

# Calculate median absolute error
med <- median(abs(predictedAge_ovary_test - y_test))

# Compute correlation
cor <- cor(predictedAge_ovary_test, y_test)

# Calculate R-squared (R²)
rsquared_value <- R2_Score(predictedAge_ovary_test, y_test)

metrics_horvarth_ovary_test <- c("ovary", "test", mae_value, med, rmse_value, rsquared_value, cor)

# Plot results
results_ovary_test <- data.frame("predicted" = predictedAge_ovary_test, "age" = y_test, "sample" =  row.names(X_test))


a <- ggplot(results_ovary_test, aes(x = age, y = predicted)) +
  geom_point() +  # Add scatter points
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +  # Add a 45-degree line
  annotate("text", x=6, y=35, label = paste("MAE =", round(mae_value, 2), "\n", "MED =", round(med, 2), "\nRMSE =", round(rmse_value, 2), "\nR2 =", round(rsquared_value, 2), "\ncor =", round(cor,2))) + 
  labs(x = "Real Age", y = "Predicted Age") +  # Label axes
   ggtitle(paste0("Real Age vs. Predicted Age (Train set, n = ", nrow(results_ovary_test),")")) +  # Add a title
  theme_classic() + 
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size = 18)) + 
  xlim(0, max(results_ovary_test$age)) +
  ylim(0, max(results_ovary_test$predicted))


print(a)

# Create a scatter plot of real age vs. predicted age
png(paste0(image_folder, "2.horvarth_ovary_test_real_vs_predicted.png"), res = 300, w = 2000, h = 1500)
a
dev.off()

```

## Prostate
 
### Load my data and data preparation
```{r}
# Load methylation data
prostate <- fread(file = "../../data/methylation_prostate.csv")

#Load metadata 
metadata <- read.csv(file = "../../metadata/eGTExDNA_Pierce_Jan18.09-11-2021.tsv", sep = "\t")
metadata <- metadata %>% filter(`Tissue.Site.Detail` == "Prostate")
age_data <- read.csv(file = "../../metadata/prostate_annotation_meth.csv")


prostate_test_samples <- read.csv(file = "../../metadata/prostate_test_metadata.csv")


#Filter for samples with age data
samples_with_age <- age_data %>% pull(Sample_ID)
prostate <- prostate %>% 
  select(probe, all_of(samples_with_age))

```

Keep only probes in the 21K (we have some missing values, which we will set to NA)

```{r}
prostate <- prostate %>% 
  filter(probe %in% probeAnnotation21kdatMethUsed$Name)


match1 <- match(probeAnnotation21kdatMethUsed$Name , prostate$probe)
sum(is.na(match1))
```

Due to missing genes from the 21K probe array, we need to input them with NA values

```{r}
missing_probes <- probeAnnotation21kdatMethUsed$Name[is.na(match1)]
missing_probes_data <- matrix(data = NA, nrow = length(missing_probes), ncol = ncol(prostate))
missing_probes_data[,1] <- missing_probes
colnames(missing_probes_data) <-  colnames(prostate)

missing_probes_data <- as.data.frame(missing_probes_data)

prostate <- rbind(prostate, missing_probes_data)
prostate <- prostate %>% 
  as_tibble() %>%
  mutate_at(2:ncol(prostate), as.numeric)
  
```

```{r}
print(dim(prostate))
```

Sanity check for missing data

```{r}
match2 <- match(probeAnnotation21kdatMethUsed$Name , prostate$probe)
sum(is.na(match2))
```

### Script based on Horvath "step-wise analysis"

#### STEP 1: DEFINE QUALITY METRICS


```{r}

meanMethBySample = as.numeric(apply(as.matrix(prostate[,-1]),2,mean,na.rm=TRUE))
minMethBySample = as.numeric(apply(as.matrix(prostate[,-1]),2,min,na.rm=TRUE))
maxMethBySample = as.numeric(apply(as.matrix(prostate[,-1]),2,max,na.rm=TRUE))

datMethUsed= t(prostate[,-1])
colnames(datMethUsed)=as.character(prostate$probe)


noMissingPerSample=apply(as.matrix(is.na(datMethUsed)),1,sum)
table(noMissingPerSample)

```

#### STEP 2: Imputing 

Run the following code if there is at least one missing

```{r}
if (max(noMissingPerSample,na.rm=TRUE)<3000 ){
  if ( max(noMissingPerSample,na.rm=TRUE)>0 ){
    dimnames1=dimnames(datMethUsed)
    datMethUsed= data.frame(t(impute.knn(t(datMethUsed))$data))
    dimnames(datMethUsed)=dimnames1
  } # end of if
} # end of if (! fastImputation )

#Note There another faster imputation method in the script


```

#### Step 2.5 Normalize the data

```{r include=FALSE}
source("normalization.R")

datMethUsedNormalized = BMIQcalibration(
  datM = datMethUsed,
  goldstandard.beta = probeAnnotation21kdatMethUsed$goldstandard2,
  plots = FALSE
)
```


#### Step 3: Split the data into train and test


```{r}
split_in_train_test <- function(meth, age_data, test_set) {
  # Metadata
  metadata_test <- age_data[age_data$tissue_sample_id %in% test_set$sample_id, ]
  metadata_train <- age_data[!age_data$tissue_sample_id %in% test_set$sample_id, ]

  # Age data
  y_test <- metadata_test$AGE
  y_train <- metadata_train$AGE

  # Methylation data
  X_train <- meth[metadata_train$Sample_ID, ]
  X_test <- meth[metadata_test$Sample_ID, ]

  return(list(X_train, X_test, y_train, y_test))
}


train_test <- split_in_train_test(datMethUsedNormalized, age_data, prostate_test_samples)

X_train <- train_test[[1]]
X_test <- train_test[[2]]

y_train <- train_test[[3]]
y_test <- train_test[[4]]

```

#### STEP 4: Predict age and create a data frame for the output 

```{r}

selectCpGsClock <- is.element(dimnames(datMethUsed)[[2]], as.character(datClock$CpGmarker[-1]))

## Train
datMethClock0 <- data.frame(X_train[,selectCpGsClock])
datMethClock <-  data.frame(datMethClock0[ as.character(datClock$CpGmarker[-1])])
predictedAge_prostate_train <- as.numeric(anti.trafo(datClock$CoefficientTraining[1]+as.matrix(datMethClock)%*% as.numeric(datClock$CoefficientTraining[-1])))

## Test
datMethClock0 <- data.frame(X_test[,selectCpGsClock])
datMethClock <-  data.frame(datMethClock0[ as.character(datClock$CpGmarker[-1])])
predictedAge_prostate_test <- as.numeric(anti.trafo(datClock$CoefficientTraining[1]+as.matrix(datMethClock)%*% as.numeric(datClock$CoefficientTraining[-1])))

```

#### Step 5: Compute accuracy metrics

In train data

```{r}

# Calculate Mean Absolute Error (MAE)
mae_value <- mae(predictedAge_prostate_train, y_train)

# Calculate Mean Squared Error (MSE)
mse_value <- mse(predictedAge_prostate_train, y_train)

# Calculate Root Mean Squared Error (RMSE)
rmse_value <- sqrt(mse_value)

# Calculate median absolute error
med <- median(abs(predictedAge_prostate_train - y_train))

# Compute correlation
cor <- cor(predictedAge_prostate_train, y_train)

# Calculate R-squared (R²)
rsquared_value <- R2_Score(predictedAge_prostate_train, y_train)

metrics_horvarth_prostate_train <- c("prostate", "train", mae_value, med, rmse_value, rsquared_value, cor)

# Plot results
results_prostate_train <- data.frame("predicted" = predictedAge_prostate_train, "age" = y_train, "sample" = row.names(X_train))


a <- ggplot(results_prostate_train, aes(x = age, y = predicted)) +
  geom_point() +  # Add scatter points
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +  # Add a 45-degree line
  #annotate(paste("MAE =", round(mae_value, 2), "\n", "RMSE =", round(rmse_value, 2))) + 
  annotate("text", x=6, y=60, label = paste("MAE =", round(mae_value, 2), "\n", "MED =", round(med, 2), "\nRMSE =", round(rmse_value, 2), "\nR2 =", round(rsquared_value, 2), "\ncor =", round(cor,2))) + 
  labs(x = "Real Age", y = "Predicted Age") +  # Label axes
   ggtitle(paste0("Real Age vs. Predicted Age (Train set, n = ", nrow(results_prostate_train),")")) +  # Add a title
  theme_classic() + 
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size = 18)) + 
  xlim(0, max(results_prostate_train$age)) +
  ylim(0, max(results_prostate_train$predicted))


print(a)

# Create a scatter plot of real age vs. predicted age
png(paste0(image_folder, "2.horvarth_prostate_train_real_vs_predicted.png"), res = 300, w = 2000, h = 1500)
a
dev.off()

```

In test data

```{r}
# Calculate Mean Absolute Error (MAE)
mae_value <- mae(predictedAge_prostate_test, y_test)

# Calculate Mean Squared Error (MSE)
mse_value <- mse(predictedAge_prostate_test, y_test)

# Calculate Root Mean Squared Error (RMSE)
rmse_value <- sqrt(mse_value)

# Calculate median absolute error
med <- median(abs(predictedAge_prostate_test - y_test))

# Compute correlation
cor <- cor(predictedAge_prostate_test, y_test)

# Calculate R-squared (R²)
rsquared_value <- R2_Score(predictedAge_prostate_test, y_test)

metrics_horvarth_prostate_test <- c("prostate", "test", mae_value, med, rmse_value, rsquared_value, cor)

# Plot results
results_prostate_test <- data.frame("predicted" = predictedAge_prostate_test, "age" = y_test, "sample" = row.names(X_test))


a <- ggplot(results_prostate_test, aes(x = age, y = predicted)) +
  geom_point() +  # Add scatter points
  geom_abline(intercept = 0, slope = 1, linetype = "dashed", color = "red") +  # Add a 45-degree line
  #annotate(paste("MAE =", round(mae_value, 2), "\n", "RMSE =", round(rmse_value, 2))) + 
  annotate("text", x=6, y=60, label = paste("MAE =", round(mae_value, 2), "\n", "MED =", round(med, 2), "\nRMSE =", round(rmse_value, 2), "\nR2 =", round(rsquared_value, 2), "\ncor =", round(cor,2))) + 
  labs(x = "Real Age", y = "Predicted Age") +  # Label axes
  ggtitle(paste0("Real Age vs. Predicted Age (Test dataset, n = ", nrow(results_prostate_test),")")) +  # Add a title
  theme_classic() + 
  theme(axis.text = element_text(size = 14), 
        axis.title = element_text(size = 18)) + 
  xlim(0, max(results_prostate_test$age)) +
  ylim(0, max(results_prostate_test$predicted))


print(a)

# Create a scatter plot of real age vs. predicted age
png(paste0(image_folder, "2.horvarth_prostate_test_real_vs_predicted.png"), res = 300, w = 2000, h = 1500)
a
dev.off()

```

Note: I have code to generate the prediction in other tissues in the version 1 folder

## Save results in tabular format
```{r}
res_horvar_clock <- data.frame("A" = metrics_horvarth_lung_train, 
                          "B" = metrics_horvarth_lung_test, 
                          "C" = metrics_horvarth_colon_train, 
                          "D" = metrics_horvarth_colon_test,
                          E = metrics_horvarth_ovary_train, 
                          "F" = metrics_horvarth_ovary_test, 
                          "G" = metrics_horvarth_prostate_train, 
                          "AA" = metrics_horvarth_prostate_test
                          )

res_horvar_clock <- t(res_horvar_clock)
res_horvar_clock <- data.frame(res_horvar_clock)
row.names(res_horvar_clock) <- NULL
colnames(res_horvar_clock) <- c("Tissue", "Dataset", "MAE", "MED", "RMSE", "R", "cor")


write.csv(res_horvar_clock, file = "../../results/2.epigenetic_clocks/Hovarth/epigenic_clocks_Horvarth.csv", quote = FALSE)
```

## Save the results from individual predictors

```{r}
write.csv(results_lung_train, "../../results/2.epigenetic_clocks/Hovarth/horvarth_lung_train_pred.csv" , quote = FALSE)
write.csv(results_lung_test, "../../results/2.epigenetic_clocks/Hovarth/horvarth_lung_test_pred.csv", quote = FALSE)

write.csv(results_colon_train, "../../results/2.epigenetic_clocks//Hovarth/horvarth_colon_train_pred.csv", quote = FALSE)
write.csv(results_colon_test, "../../results/2.epigenetic_clocks/Hovarth/horvarth_colon_test_pred.csv", quote = FALSE)

write.csv(results_ovary_train, "../../results/2.epigenetic_clocks/Hovarth/horvarth_ovary_train_pred.csv", quote = FALSE)
write.csv(results_ovary_test, "../../results/2.epigenetic_clocks/Hovarth/horvarth_ovary_test_pred.csv", quote = FALSE)

write.csv(results_prostate_train, "../../results/2.epigenetic_clocks/Hovarth/horvarth_prostate_train_pred.csv", quote = FALSE)
write.csv(results_prostate_test, "../../results/2.epigenetic_clocks/Hovarth/horvarth_prostate_test_pred.csv", quote = FALSE)
```

